---
title: "Computing for the Social Sciences: Lecture 10"
subtitle: "Topics: Exam 1; Dates with lubridate"
author: "Sabrina Nardin, Fall 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---


# Agenda {.center}

::: {.agenda-list}
1. Exam 1
2. Manipulating Dates with lubridate  
:::


### Slides are in progress and will be updated before class.

::: {.content-hidden}


<!--
FALL 26
ADD INTRO TO EXAM 1 AND LOGISTICS HERE 
-->


# 1. Exam 1


# 2. Manipulating Dates with Lubridate {.slide .center .middle}

---

## Date and Times with Lubridate

Lubridate (also part of the `tidyverse`) is a handy package that makes it easier work with dates & times <https://r4ds.hadley.nz/datetimes.html>

::: {.fragment}

### Common Lubridate functions:

- Turn strings into dates or date-times with `ymd()`, `mdy()`, `dmy()`, `ymd_hms()`  
- Turn variables into dates or date-times with `make_datetime()`  
- Extract individual date components with `year()`, `month()`, `day()`, `wday()`  
- Round the date to a nearby unit of time `+ days(1)`, `floor_date()`, `round_date*()`  
- Handle time spans (multiple options, see Ch. 17)
- Handle time zones `with_tz()`, `force_tz()`  
:::

---

## Turn strings into dates or date-times

```{r, echo = TRUE}
library(lubridate)
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")
ymd_hms("2017-01-31 20:11:59")
```

---

## Turn variables into dates or date-times

::: {.columns}
::: {.column width="40%"}
```{r, echo = TRUE}
library(tidyverse)
library(nycflights13)
flights |> 
  select(year, month, day, hour, minute)
```
:::

::: {.column width="60%"}
```{r, echo = TRUE}
# create a date/time from this data with make_datetime()
flights |> 
  select(year, month, day, hour, minute) |> 
  mutate(departure = make_datetime(year, month, day, hour, minute))

```
:::
:::

<br>

More examples: <https://r4ds.hadley.nz/datetimes.html#from-individual-components>

</br>

---

## Extract individual date components 

```{r, echo = TRUE}
datetime <- ymd_hms("2025-10-22 12:30:33")

year(datetime)
month(datetime)
mday(datetime)
yday(datetime)
wday(datetime)
```

More examples: <https://r4ds.hadley.nz/datetimes.html#getting-components>

---

##  Round the date to a nearby unit of time 

```{r, echo = TRUE}
# create date object from string
dt <- ymd_hms("2025-10-22 12:45:00")
dt

# round down with floor_date() 
floor_date(dt, "day")
floor_date(dt, "month")

# round up with ceiling_date()
ceiling_date(dt, "day")
ceiling_date(dt, "month")
```

More examples: <https://r4ds.hadley.nz/datetimes.html#rounding>

---

## Handle Time Spans

```{r, echo = TRUE}
# How old are you?
age <- today() - ymd("2001-10-22")
age

as.duration(age)
```

More examples: <https://r4ds.hadley.nz/datetimes.html#time-spans>

---

## Handle Time Zones

```{r, echo = TRUE}
greenwitch <- ymd_hms("2025-10-22 6:45:00")

# display same instant in another time zone
# shows local Central Time equivalent
with_tz(greenwitch, tzone = "America/Chicago")

# assign a time zone (without shifting actual clock time)
# interprets existing time as if it were recorded in Chicago
force_tz(greenwitch, "America/Chicago")
```

More examples: <https://r4ds.hadley.nz/datetimes.html#time-zones>
